---
title: "Supplementary tables"
author: "Pietro Demela, Nicola Pirastu, Blagoje Soskic"
date: "11/01/2023"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
knitr::opts_knit$set(root.dir =  "/project/aid_sharing/AID_sharing") #set up the project directory
```

```{r}
library(data.table)
library(stringr)
library(dplyr)
```



## Supplementary table S2, LDSC SE and pvalues

```{r paged.print=FALSE}

#load the output of ldsc 

LDSCoutput <- readRDS('outputs/rev_1/02_munge_ldsc_sumstats/ldsc_output_rev1.RDS')

	k<-nrow(LDSCoutput$S)
	SE<-matrix(0, k, k)
	SE[lower.tri(SE,diag=TRUE)] <-sqrt(diag(LDSCoutput$V))
	
	#add rownames
	colnames(SE) <- colnames(LDSCoutput$S)
  row.names(SE) <- 	colnames(SE) 
	
	#obtain the p-values
	pvalues <- 2*pnorm(abs(LDSCoutput$S/SE),lower.tail=FALSE)
	rownames(pvalues) <- colnames(pvalues)
	
	#print pvalues
	signif(pvalues,3)
	#print SE
	signif(SE,3)
	
	write.csv(signif(pvalues,3),file = 'outputs/rev_1/Supplementary_tables/2_Supplementary_table_pvalues_ldsc.csv', row.names = T, col.names = T, quote = F)
	write.csv(signif(SE,3),file = 'outputs/rev_1/Supplementary_tables/2_Supplementary_table_se_ldsc.csv', row.names = T, col.names = T, quote = F)

```


```{r}
#load the output of ldsc 

LDSCoutput <- readRDS('outputs/rev_1/02_munge_ldsc_sumstats/ldsc_output_rev1.RDS')


	#add rownames
	row.names(LDSCoutput$S) <- colnames(LDSCoutput$S)
	#print pvalues
	varcov <- signif(LDSCoutput$S,3)
	#print SE
	
	
	write.csv(varcov,file = 'outputs/rev_1/Supplementary_tables/2_Supplementary_table_varcovar_ldsc.csv', row.names = T, col.names = T, quote = F)
	

```


## Supplementary table S3, genomic regions Factors

```{r paged.print=FALSE}

#----Supplementary table 1, table of genomic regions FACTORS -------------------
loci_factors <- fread('outputs/rev_1/04_genomic_regions_factors/regions_factors_with_index.txt', data.table = F) 
head(loci_factors)

#change name to the factor name
loci_factors$trait[which(loci_factors$trait=='f1')] <- 'Fgut'
loci_factors$trait[which(loci_factors$trait=='f2')] <- 'Faid'
loci_factors$trait[which(loci_factors$trait=='f3')] <- 'Falrg'

#rename the columns (here the effect Allele is A1)
gen_reg <- loci_factors %>% rename('p-value'=P , Factor ='trait',  'genomic_region(chr_start_end)'= pan_locus_name, 'lead_SNP'=SNP, 'OTHA'=A1, 'REFA'=A2) %>% select(-c( pan_locus)) 
colnames(gen_reg) <- toupper(colnames(gen_reg))

#save
fwrite(gen_reg, 'outputs/rev_1/Supplementary_tables/3_Supplementary_table_genomic_regions_factors.csv', sep = ',', col.names = T, quote = F)

```


## Supplementary table S4, conditional loci Factors and v2g genes

```{r paged.print=FALSE}

loci.table <-  fread('outputs/rev_1/10_pathways/factor_loci_genes.txt', data.table = F) 
loci.table$final.locus=paste0(loci.table$Chr,":",loci.table$start,"-",loci.table$end,"_",loci.table$sub_locus)

#Add the column for independent signals from the same trait ended up colocalising in the same locus
loci.table$indip_coloc <- NA

for(i in 1:3){
  tt <- c('f1', 'f2', 'f3')[i]
  dup <- loci.table[which(loci.table$trait== tt), ]$final.locus[duplicated(loci.table[which(loci.table$trait==tt), ]$final.locus)]
  loci.table[which(loci.table$trait== tt & loci.table$final.locus %in% dup), ]$indip_coloc <- 'multiple singal colocalising in the same locus'
  loci.table[which(loci.table$trait== tt & (!loci.table$final.locus %in% dup)), ]$indip_coloc <- '-'
}


#change the factor names
loci.table$trait[which(loci.table$trait=='f1')] <- 'Fgut'
loci.table$trait[which(loci.table$trait=='f2')] <- 'Faid'
loci.table$trait[which(loci.table$trait=='f3')] <- 'Falrg'

#renmae the columns
loci <- loci.table %>% rename(Factor=trait, 'V2G_Opentrgets_EnsemblID'= v2g, 
                              'V2G_Opentrgets_gene_name'=v2g_gene_name, 'LOCUS_NAME(CHR_START_END)'= final.locus, 'Beta'=b, 'BetaJ'=bJ, 'NOTE'=indip_coloc  ) %>% 
  select(-c(pan.locus, sub_locus) )
colnames(loci) <- toupper(colnames(loci))

head(loci)

#save
fwrite(loci,'outputs/rev_1/Supplementary_tables/4_Supplementary_table_conditional_loci_withv2g_factors.csv', sep = ',', col.names = T, quote = F)


```


## Supplementary table S5, colocalization results traits

```{r message=TRUE, paged.print=FALSE}

posterior <- fread('outputs/rev_1/coloc_eqtl_mr/loci_definitions/colocalization.table.all.tsv', data.table = F)
regions <- fread('outputs/rev_1/06_genomic_regions_all_traits_munged/loci_all_traits.txt')

posterior$genomic_region <- '-'

for(i in 1:nrow(posterior)){
  
  posterior[i,]$genomic_region <- unique(regions[which(regions$pan_locus== posterior[i,]$locus), ]$pan_locus_name)

  }


final_post <- posterior %>% select(-c('locus')) %>% rename('genomic_region(CHR_START_END)'=genomic_region, 'trait_1'=t1, 'trait_2'=t2, 'leadSNP_locus_trait_1'=hit1, 'leadSNP_locus_trait_2'=hit2)

final_post$trait_1[which(final_post$trait_1=='f1')] <- 'Fgut'
final_post$trait_2[which(final_post$trait_2=='f1')] <- 'Fgut'

final_post$trait_1[which(final_post$trait_1=='f2')] <- 'Faid'
final_post$trait_2[which(final_post$trait_2=='f2')] <- 'Faid'

final_post$trait_1[which(final_post$trait_1=='f3')] <- 'Falrg'
final_post$trait_2[which(final_post$trait_2=='f3')] <- 'Falrg'

final_post$trait_1[which(final_post$trait_1=='derma')] <- 'eczema'
final_post$trait_2[which(final_post$trait_2=='derma')] <- 'eczema'


fwrite(final_post,'outputs/rev_1/Supplementary_tables/5_Supplementary_table_coloc_all_loci_results.csv', sep = ',', col.names = T, quote = F)

head(final_post)






```


## Supplementary table S6, genomic regions all traits and factors

```{r paged.print=FALSE}
loci_factors <- fread('outputs/rev_1/06_genomic_regions_all_traits_munged/loci_all_traits.txt', data.table = F) 
head(loci_factors)

#change name to the factor name
loci_factors$trait[which(loci_factors$trait=='f1')] <- 'Fgut'
loci_factors$trait[which(loci_factors$trait=='f2')] <- 'Faid'
loci_factors$trait[which(loci_factors$trait=='f3')] <- 'Falrg'
loci_factors$trait[which(loci_factors$trait=='derma')] <- 'eczema'

#rename the columns (here the effect allele is A2)
gen_reg <- loci_factors %>% rename('p-value'=P , Factor ='trait',  'genomic_region(chr_start_end)'= pan_locus_name, 'lead_SNP'=SNP, 'OTHA'=A2, 'REFA'=A1) %>% select(-c( pan_locus)) 
colnames(gen_reg) <- toupper(colnames(gen_reg))

head(gen_reg)
#save
fwrite(gen_reg, 'outputs/rev_1/Supplementary_tables/6_Supplementary_table_genomic_regions_all_traits.csv', sep = ',', col.names = T, quote = F)

```

## Supplementary table S7, conditional loci traits and factors

```{r paged.print=FALSE}
# all_loci <- fread('outputs/rev_1/Supplementary_tables/7_Supplementary_all_conditional_loci.csv', data.table = F)
# 
# all_loci[which(all_loci$TRAIT=='derma'),]$TRAIT <- 'eczema'
# head(all_loci)
# dim(all_loci)
# 
# fwrite(gen_reg, 'outputs/rev_1/Supplementary_tables/7_Supplementary_all_conditional_loci.csv', sep = ',', col.names = T, quote = F)
```

## Supplementary table S8, KEGG pathways

### load mart

```{r}
library(biomaRt)
mart <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart)
```


```{r paged.print=FALSE}
kegg <-  readRDS('outputs/rev_1/10_pathways/kegg.RDS')

head(kegg$result)

#--------- kegg table complete -------------------------------------------------
kg <- kegg$result[, c('query','p_value', 'term_name', 'intersection')] 

#retrieve the genbe symbols
he <- kegg$result
genes <- unique(strsplit(paste0(he$intersection, collapse = ','), split = ',')[[1]])
genes_symbol <- getBM(filters= "ensembl_gene_id", attributes= c('ensembl_gene_id', 'hgnc_symbol'),values=  genes,mart=mart )

head(genes_symbol)

#add a gene symbol column
kg$hgnc <- rep('NA', nrow(kg))
for(i in 1:nrow(kg)){
  kg$hgnc[i] <- paste0(genes_symbol[ which(genes_symbol$ensembl_gene_id %in% strsplit( kg$intersection[i], split = ',')[[1]]) ,]$hgnc_symbol, collapse = ', ')
}

head(kg)


#format for exporting with the rigth names
colnames(kg) <- c('Trait', 'P-value (adjusted)', 'Pathway', 'ENSEMBL', 'Gene symbols')
kg$Trait <- toupper(kg$Trait)
kegg.table <- kg[,c(3,1,2, 5)]


#renmae the factors
kegg.table[which(kegg.table$Trait=='F1'),]$Trait <- 'Fgut'
kegg.table[which(kegg.table$Trait=='F2'),]$Trait <- 'Faid'
kegg.table[which(kegg.table$Trait=='F3'),]$Trait <- 'Falrg'

#round the pvalues
kegg.table$`P-value (adjusted)` <- signif(kegg.table$`P-value (adjusted)`,4)

#save the csv
write.table(kegg.table, sep = ',', file = 'outputs/rev_1/Supplementary_tables/8_Supplementary_kegg.table.csv', row.names = F)

```

## Supplementary table S9, GO terms

```{r paged.print=FALSE}

#go table complete, (all the results)
go <-  readRDS('outputs/rev_1/10_pathways/go.RDS')
gg <-  go$result[, c('query','p_value', 'term_name', 'intersection')] 

genes <- unique(strsplit(paste0(gg$intersection, collapse = ','), split = ',')[[1]])
genes_symbol <- getBM(filters= "ensembl_gene_id", attributes= c("entrezgene_id","ensembl_gene_id", 'hgnc_symbol'),values=  genes,mart=mart )
gg$hgnc <- rep('NA', nrow(gg))

for(i in 1:nrow(gg)){
  gg$hgnc[i] <- paste0(genes_symbol[ which(genes_symbol$ensembl_gene_id %in% strsplit( gg$intersection[i], split = ',')[[1]]) ,]$hgnc_symbol, collapse = ',  ')
}
#add the names
colnames(gg) <- c('Trait', 'P-value (adjusted)', 'GO term', 'ENSEMBL', 'Gene symbols')
gg$Trait <- toupper(gg$Trait)
go.table <- gg[,c(3,1,2, 5,4)]

head(go.table)



#renmae the factors
go.table[which(go.table$Trait=='F1'),]$Trait <- 'Fgut'
go.table[which(go.table$Trait=='F2'),]$Trait <- 'Faid'
go.table[which(go.table$Trait=='F3'),]$Trait <- 'Falrg'

#round the pvalues
go.table$`P-value (adjusted)` <- signif(go.table$`P-value (adjusted)`,4)


#save the table ascsv
write.table(go.table, sep = ',', file = 'outputs/rev_1/Supplementary_tables/9_Supplementary_go.table.csv', row.names = F)


```

## Supplementary table S10, eQTL colocalization results

```{r}
qtl_names <- list.files('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/')[which(str_ends(list.files('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/'), 'coloc.tsv'))]


qtl_l <- list()
qtl_ok <- list()

for(i in qtl_names){
  
  qtl_l[[i]] <- try(fread(paste0('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/', i), data.table=F))
  
  if(c(!is(qtl_l[[i]], 'try-error') & !ncol(qtl_l[[i]])==1)){
    qtl_ok[[i]] <- qtl_l[[i]]
  }
  
}
```


```{r paged.print=FALSE}

qtl_res<- Reduce(rbind, qtl_ok)
qtl_factors <- qtl_res[which(qtl_res$t1 %in% c('f1', 'f2', 'f3')),]


qtl_final <- qtl_factors %>% rename('hit_trait_1'=hit1, 'hit_trait_2'=hit2, 'cell_type'=cell_type, 'trait_1'= t1, 'trait_2'=t2  )

qtl_final$trait_1[which(qtl_final$trait_1=='f1')] <- 'Fgut'
qtl_final$trait_1[which(qtl_final$trait_1=='f2')] <- 'Faid'
qtl_final$trait_1[which(qtl_final$trait_1=='f3')] <- 'Falrg'

colnames(qtl_final) <- toupper(colnames(qtl_final))

head(qtl_final)
dim(qtl_final)

#save
fwrite(qtl_final,'outputs/rev_1/Supplementary_tables/10_Supplementary_table_eQTL_results.csv', sep = ',', col.names = T, quote = F)


```
## Supplementary table S11, pQTL coloc results

```{r message=TRUE, warning=FALSE}
pqtl_list <- list.files('outputs/rev_1/pqtls')[which(str_ends(list.files('outputs/rev_1/pqtls'), 'coloc.tsv'))]

pqtl_l <- list()
pqtl_ok <- list()

#load the files that were split by chr, and cell type
for(i in pqtl_list){
  
  pqtl_l[[i]] <- try(fread(paste0('outputs/rev_1/pqtls/',i), data.table=F))
  
  if(!is(pqtl_l[[i]], 'try-error')){
    pqtl_ok[[i]] <- pqtl_l[[i]]
  }
  
}
```


```{r}
#add locus name to pqtl ias column
for(i in 1:length(names(pqtl_ok))){
   a <- str_split(names(pqtl_ok)[i],'_pqtl_coloc.tsv')[[1]][[1]]
   b <- str_split(a, 'Locus_')[[1]][[2]]
   pqtl_ok[[names(pqtl_ok)[i]]]$locus <- b
}

#put the pQTL coloc result together 
pqtl_res <- Reduce(rbind, pqtl_ok)
pqtl_res <- pqtl_res[which(pqtl_res$t1 %in% c('f1', 'f2', 'f3')),]

head(pqtl_res)
dim(pqtl_res)
#add the locus name
#add the locus name to the table
loci <- fread('outputs/rev_1/coloc_eqtl_mr//loci_definitions/final_locus_table.tsv', data.table = F)
loci$final.locus=paste0(loci$Chr,":",loci$start,"-",loci$end,"_",loci$sub_locus)
loci$tmp <- paste0(loci$pan.locus, '_',loci$sub_locus)
head(loci)

#add the locus name in this loop 
pqtl_res$locus_name <- rep('', nrow(pqtl_res))
for(i in 1:nrow(pqtl_res)){
  pqtl_res[i, ]$locus_name <- loci[which(loci$tmp==pqtl_res[i, ]$locus),]$final.locus[1]
  
}

pqtl_final <- pqtl_res %>% rename( 'locus_name_(chr:start_end)'=locus_name) %>% dplyr::select(-c('locus'))
pqtl_final <- pqtl_final[which(pqtl_final$t1 %in% c('f1', 'f2', 'f3')), ]
dim(pqtl_final)

#rename factors 
pqtl_final$t1[which(pqtl_final$t1=='f1')] <- 'Fgut'
pqtl_final$t1[which(pqtl_final$t1=='f2')] <- 'Faid'
pqtl_final$t1[which(pqtl_final$t1=='f3')] <- 'Falrg'

names(pqtl_final) <- toupper(names(pqtl_final))


head(pqtl_final)
dim(pqtl_final)

#save
fwrite(pqtl_final,'outputs/rev_1/Supplementary_tables/11_Supplementary_table_pQTL_coloc_results.csv', sep = ',', col.names = T, quote = F)


```


## Supplementary table S12, pQTL MR results

```{r paged.print=TRUE}
pqtl_mr_list <- list.files('outputs/rev_1/pqtls')[which(str_ends(list.files('outputs/rev_1/pqtls'), 'MR.tsv'))]

pmr_l <- list()
pmr_ok <- list()

#load the files that were split by chr, and cell type
for(i in pqtl_mr_list){
  
  pmr_l[[i]] <- try(fread(paste0('outputs/rev_1/pqtls/',i), data.table=F))
  
  if(!is(pmr_l[[i]], 'try-error')){
    pmr_ok[[i]] <- pmr_l[[i]]
  }
  
}
```

```{r}
#put the pQTL coloc result together 
pmr_res <- Reduce(rbind, pmr_ok)
pmr_res <- pmr_res[which(pmr_res$trait %in% c('f1', 'f2', 'f3')),]

head(pmr_res)
dim(pmr_res)
#add the locus name
#add the locus name to the table
loci <- fread('outputs/rev_1/coloc_eqtl_mr/loci_definitions/final_locus_table.tsv', data.table = F)
loci$final.locus=paste0(loci$Chr,":",loci$start,"-",loci$end,"_",loci$sub_locus)
loci$tmp <- paste0(loci$pan.locus, '_',loci$sub_locus)
head(loci)

#add the locus name in this loop 
pmr_res$locus_name <- rep('', nrow(pmr_res))
for(i in 1:nrow(pmr_res)){
  pmr_res[i, ]$locus_name <- loci[which(loci$tmp==pmr_res[i, ]$locus),]$final.locus[1]
  
}

pmr_final <- pmr_res %>% rename( 'locus_name_(chr:start_end)'=locus_name) %>% dplyr::select(-c('locus'))
pmr_final <- pmr_final[which(pmr_final$trait %in% c('f1', 'f2', 'f3')), ]
dim(pmr_final)

#rename factors 
pmr_final$trait[which(pmr_final$trait=='f1')] <- 'Fgut'
pmr_final$trait[which(pmr_final$trait=='f2')] <- 'Faid'
pmr_final$trait[which(pmr_final$trait=='f3')] <- 'Falrg'

names(pmr_final) <- toupper(names(pmr_final))
pmr_final <- pmr_final[which(!is.na(pmr_final$B)),]

head(pmr_final)
dim(pmr_final)

#save
fwrite(pmr_final,'outputs/rev_1/Supplementary_tables/12_Supplementary_table_pQTL_MR_results.csv', sep = ',', col.names = T, quote = F)

```

## Supplementary table S13 eQTL MR supplementary table

```{r paged.print=TRUE}
#mr table
mr_list <- list.files('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/')[which(str_ends(list.files('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/'), 'MR.tsv'))]

mr_l <- list()
mr_ok <- list()

#load the files that were split by chr, and cell type
for(i in mr_list){
  
  mr_l[[i]] <- try(fread(paste0('outputs/rev_1/coloc_eqtl_mr/sc_eqtl/',i), data.table=F))
  
  if(!is(mr_l[[i]], 'try-error')){
    mr_ok[[i]] <- mr_l[[i]]
  }
  
}

```


```{r paged.print=FALSE}

mr_res <- Reduce(rbind, mr_ok)
mr_res <- mr_res[which(mr_res$trait %in% c('f1', 'f2', 'f3')),]

head(mr_res)

#add the locus name
#add the locus name to the table
loci <- fread('outputs/rev_1/coloc_eqtl_mr/loci_definitions/final_locus_table.tsv', data.table = F)
loci$final.locus=paste0(loci$Chr,":",loci$start,"-",loci$end,"_",loci$sub_locus)
loci$tmp <- paste0(loci$pan.locus, '_',loci$sub_locus)
head(loci)

#add the locus name in this loop 
mr_res$locus_name <- rep('', nrow(mr_res))
for(i in 1:nrow(mr_res)){
  mr_res[i, ]$locus_name <- loci[which(loci$tmp==mr_res[i, ]$locus),]$final.locus[1]
  
}

mr_final <- mr_res %>% rename( 'locus_name_(chr:start_end)'=locus_name, 'cell_type'=cell.type) %>% dplyr::select(-c('locus'))

#rename factors 
mr_final$trait[which(mr_final$trait=='f1')] <- 'Fgut'
mr_final$trait[which(mr_final$trait=='f2')] <- 'Faid'
mr_final$trait[which(mr_final$trait=='f3')] <- 'Falrg'

names(mr_final) <- toupper(names(mr_final))

#remove duplicated rows for multiple locus associated with the same eQTL
mr_final <- mr_final[which(!is.na(mr_final$B)),]

head(mr_final)
dim(mr_final)

#save
fwrite(mr_final,'outputs/rev_1/Supplementary_tables/13_Supplementary_table_eQTL_MR_results.csv', sep = ',', col.names = T, quote = F)


```

```{r}
sessionInfo()
```

