---
title: "Figure 3"
output: 
  html_document: 
    toc: yes
author: 'Pietro Demela, Nicola Pirastu, Blagoje Soskic'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
knitr::opts_knit$set(root.dir =  "/project/aid_sharing/AID_sharing") #set up the project directory
```

## Libraries

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(data.table)
library(stringr)
library(biomaRt)
library(ComplexHeatmap)
library(RColorBrewer)
library(dplyr)
```

## Figure 3A

### Load dataset and mart

Mart for gene name conversion
```{r}

kegg <- readRDS('outputs/rev_1/10_pathways/kegg.RDS')
mart <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart)

```

### Get gene names from ensemble gene ID

```{r paged.print=FALSE}
#extract the genes that were into a kegg pathway
res <- kegg$result
genes <- unique(strsplit(paste0(res$intersection, collapse = ','), split = ',')[[1]])
head(genes)
#get the gene names from ensmebl gene id 
genes_symbol <- getBM(filters= "ensembl_gene_id", attributes= c('ensembl_gene_id', 'hgnc_symbol'),values=  genes,mart=mart )

head(genes_symbol)

#check for duplicated entries
table(duplicated(genes_symbol$ensembl_gene_id))
table(duplicated(genes_symbol$hgnc_symbol))

t_t <- data.frame(genes) 


#create a mapping table between ensmeble and geneID ordered as the gene lists that follows the order of factors in kegg
colnames(genes_symbol)[1] <- 'genes'
dim(t_t)
t_t <- inner_join(t_t, genes_symbol, 'genes')
head(t_t)



```

Create a matrix containing the unique pathway name and the columns for each factor that will serve to plot the -log10pvalues

```{r}

kg <- kegg$result[,c('query', 'p_value', 'term_name')]

#create mtp, a matrix containing the unique pathway name and the columns containing the factors
mtp <- matrix(nrow = length(unique(kg$term_name)), ncol = 3)
colnames(mtp) <- c('f1', 'f2', 'f3')
rownames(mtp) <- unique(kg$term_name)

#show mtp
mtp

#populate mtp with the z scores
for(i in c('f1', 'f2', 'f3')){
      kg[ which(kg$query==i),]$term_name
      mtp[, i][base::match(  kg[ which(kg$query==i), ]$term_name, names(mtp[, i]))] <- -log10(kg[ which(kg$query==i), ]$p_value) #important to set the radius of the circle
}

#show populated mtp
mtp

# add zeros to the empty rows
for(i in c('f1', 'f2', 'f3')){
  mtp[,i][which(is.na(mtp[,i]))] <- 0
}

#show mtp
mtp

#toupper case
colnames(mtp) <- toupper(colnames(mtp))

#order mtp with the shared pathways on top and the genes to be grouped by factors
otp <- mtp[c(1,6,2,4,7,8,3,5,9,10,11,12),]

otp
```


Create a table and add the information about the genes, shared or not.
gpt table stores the information about a certain gene belonging to a certain pathway

```{r}
#gtp matrix will store the information about a certain gene belonging to a certain pathway
gtp <- matrix(nrow = length(unique(kg$term_name)), ncol = nrow(genes_symbol))
rownames(gtp) <- unique(kg$term_name)
colnames(gtp) <- genes_symbol$hgnc_symbol

#it looks like that
gtp



#populate the matrix
for(k in 1:length(unique(res$term_name)) ){
  
        tt <- unique(res$term_name)[k]
        g_s <- list()
        u <- vector()
        
        #if the intersection is between three pathways
        if( table(res$term_name)[tt] == 3 ){
          
          
                  for(j in 1:3){
                           
                     
                          u <-  res[which(res$term_name==tt),][j,]$query
                          g_s[[u]] <-  genes_symbol[which(genes_symbol$genes %in% str_split( res[which(res$term_name==tt),][j,]$intersection, pattern = ','  )[[1]]) ,]$hgnc_symbol
                         
                          
                  }        
                
                    #make a combination matrix and search for all the combinations
                    cmb <- make_comb_mat(g_s, mode = 'distinct')
                    
                    if('110' %in% comb_name(cmb)){
                      gtp[tt,extract_comb(cmb, '110')] <- paste0(names(g_s)[c(1,2)], collapse = '-') 
                    }
                    
                    if('011' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '011')] <- paste0(names(g_s)[c(2,3)], collapse = '-') 
                    }
                    
                     if('101' %in% comb_name(cmb)){
                      gtp[tt,extract_comb(cmb, '101')] <- paste0(names(g_s)[c(1,3)], collapse = '-') 
                    }
                    
                    if('100' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '100')] <- paste0(names(g_s)[1], collapse = '-')
                    }
                    
                    if('010' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '010')] <- paste0(names(g_s)[2], collapse = '-')
                    }
                    
                    if('001' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '001')] <- paste0(names(g_s)[3], collapse = '-')
                    
                    }
                    
                    if('111 '%in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '111')] <- paste0(names(g_s)[1:3], collapse = '-')
                    
                    }
                
          rm(cmb)
          
            #if the intersection is between two elements do this
        } else if( table(res$term_name)[tt] == 2 ){
           
              for(i in 1:nrow(res[which(res$term_name==tt),])){
                           
                     
                          u <-  res[which(res$term_name==tt),][i,]$query
                          g_s[[u]] <-  genes_symbol[which(genes_symbol$genes %in% str_split( res[which(res$term_name==tt),][i,]$intersection, pattern = ','  )[[1]]) ,]$hgnc_symbol
                         
                          
                  }        
                
                    #make a combination matrix and search for all the combinations
                   cmb <- make_comb_mat(g_s, mode = 'distinct')
                    
                    if('11' %in% comb_name(cmb)){
                      gtp[tt,extract_comb(cmb, '11')] <- paste0(names(g_s)[1:2], collapse = '-') 
                    }
                    
                    if('01' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '01')] <- paste0(names(g_s)[2], collapse = '-') 
                    }
                    
                    if('10' %in% comb_name(cmb)){
                    gtp[tt,extract_comb(cmb, '10')] <- paste0(names(g_s)[1], collapse = '-')
                    }
                    
          rm(cmb)
          
         }
           
           
          #if there is no intersection proceed as follows
          else if( table(res$term_name)[tt] == 1 ){
          
            for(i in 1:nrow(res[which(res$term_name==tt),])){
                      ges <-  genes_symbol[which(genes_symbol$genes %in% str_split( res[which(res$term_name==tt),][i,]$intersection, pattern = ','  )[[1]]) ,]$hgnc_symbol
                      gtp[tt ,ges] <- res[which(res$term_name==tt),][i,]$query
          }
        }
        
        
}



ggtp <-  gtp[rownames(otp),t_t$hgnc_symbol]

#look at the combinations
names(table(ggtp))

#show ggtp
ggtp
```



### Plot heatmap section

```{r}
colori <- ComplexHeatmap:::default_col(gtp[,t_t$hgnc_symbol])
colori[c('f1', 'f2', 'f3', 'f1-f3', 'f2-f3', 'f1-f2')] <-  rep('white', 6)

a<-Heatmap(ggtp, 
           #column_title = "Genes and pathways KEGG",
           rect_gp = gpar(col = brewer.pal(5, 'Greys')[2], lwd = 0.25), 
           column_title_gp = gpar(fontsize = 20, fontface = "bold"),
           col=colori, 
           border=T,
           row_names_gp = gpar( fontsize=0),
           column_names_gp = gpar(fontsize=25),
           column_names_side = 'bottom',
           show_heatmap_legend = F,
           row_title = "Pathways",
           na_col = 'white', 
           height = unit(6.5, 'cm'), width = 28, 
           top_annotation =  columnAnnotation(legend = anno_empty(border = F, width =  unit(28, "cm"), height = unit(2, 'cm'))), #make space for the annotation
           
           
           cell_fun = function(j, i, x, y, width, height, fill) {
             if(!is.na(ggtp[,t_t$hgnc_symbol][i, j] )){
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f1' ){
                 grid.points(x = x, y = y, pch = 22, size = unit(18, "mm"), gp = gpar(col = brewer.pal(6, 'Paired')[1], fill=brewer.pal(5, 'Paired')[2] ))
               }
               
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f2' ){
                 grid.points(x = x, y = y, pch = 22, size = unit(18, "mm"), gp = gpar(col = brewer.pal(5, 'Paired')[3], fill=brewer.pal(5, 'Paired')[4]))
               }
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f3' ){
                 grid.points(x = x, y = y, pch = 22, size = unit(18, "mm"), gp = gpar(col = brewer.pal(6, 'Paired')[5], fill=brewer.pal(6, 'Paired')[6]))
               }
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f1-f3' ){
                 grid.points(x = x, y = y, pch=22, size = unit(18, "mm"), gp = gpar(col = 'grey', fill='black'))
               }
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f2-f3' ){
                 grid.points(x = x, y = y, pch=22, size = unit(18, "mm"), gp = gpar(col = 'grey', fill='yellow'))
               }
               
               if(ggtp[,t_t$hgnc_symbol][i, j]== 'f1-f2' ){
                 grid.points(x = x, y = y, pch=22, size = unit(18, "mm"), gp = gpar(col = 'black', fill='grey'))
               }
               
             }
             
           }
)
```


### Plot the pvalues section

```{r}
b <- Heatmap(otp,  
             rect_gp = gpar(col = brewer.pal(5, 'Greys')[2], lwd = 0.25), 
             cluster_rows = F,
             cluster_columns = F,
             col = c('white', 'white'), 
             border=T,
             show_column_dend = F,
             show_row_dend = F,
             row_names_gp = gpar(fontsize=3),
             row_names_side = 'left',
             width = unit(6.5, 'cm'),
             height = unit(28, 'cm'),
             show_heatmap_legend = F,
             column_names_gp = gpar(fontsize=25),
             column_names_side = 'bottom',
             column_names_rot = 360,
             top_annotation =  columnAnnotation(legend = anno_empty(border = F,
                                                                    width =  unit(28, "cm"), height = unit(2, 'cm'))), #create an empty annotation for adding the legend elements 
             
             cell_fun = function(j, i, x, y, width, height, fill) {
               if(otp[i, j] > 0){
                 grid.circle(x = x, y = y, r = unit(otp[i,j], 'mm'), gp = gpar(fill = brewer.pal(7,'BrBG')[3], col = 'black')) #the unit of the -log10(pvalues) is mm as in the legend
                 #grid.text(signif((10^100)^(-otp[i,j]),2), x=x,  y = y, gp=gpar(fontsize=10)) #in case you wish to see the pvalues in the center of the circle
                 
               }
             }
)

```


### Save as pdf

```{r}
#save as pdf
pdf('outputs/rev_1/Figures/Fig_3/3A_semaphor_plot.pdf', height = 36, width = 36)
#plot the heatmaps
b+a
#for adding the legend on the pvalues scales, create the same heatmap with fixed values and copy paste them in illustrator 
for(k in 1:6) {
  i <- c(0.05,0.30,0.50, 0.70, 0.80,0.90)[k] #this are the positions 
  
  decorate_annotation('legend', {
    
    grid.circle(x=i, r= unit(-log10(c(10^-10,10^-8,10^-6,10^-4,10^-2,0.05)), 'mm')[k],gp = gpar(fill = brewer.pal(7,'BrBG')[3], col = 'black')) #this are the pvalues that are shown in the legend, the dimension of the -log10(p) is in mm
    grid.text(x=i, round(-log10(c(10^-10,10^-8,10^-6,10^-4,10^-2,0.05)), 3)[k]) #show the p
  })
}

dev.off()
```

### Plot it here

```{r fig.height=30, fig.width=30}

b+a
#for adding the legend on the pvalues scales, create the same heatmap with fixed values and copy paste them in illustrator 
for(k in 1:6) {
  i <- c(0.05,0.30,0.50, 0.70, 0.80,0.90)[k] #this are the positions 
  
  decorate_annotation('legend', {
    
    grid.circle(x=i, r= unit(-log10(c(10^-10,10^-8,10^-6,10^-4,10^-2,0.05)), 'mm')[k],gp = gpar(fill = brewer.pal(7,'BrBG')[3], col = 'black')) #this are the pvalues that are shown in the legend, the dimension of the -log10(p) is in mm
    grid.text(x=i, round(-log10(c(10^-10,10^-8,10^-6,10^-4,10^-2,0.05)), 3)[k]) #show the p
  })
}

```


```{r}
sessionInfo()
```

















