---
title: "07 Q heterogeneity index"
author: "Pietro Demela"
date: "10/01/2023"
output: html_document
---
In order to evaluate whether the calculated SNP effects were acting through our three factor model, we performed the QSNP heterogeneity tests. The heterogeneity test returns a 𝜒2, whose null hypothesis suggests that the SNP is acting through the specified model. Therefore, rejecting the null hypothesis means that the SNP acts through a model that is different from the specified one.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
knitr::opts_knit$set(root.dir =  "/project/aid_sharing/AID_sharing") #set up the project directory
```

## Libraries

```{r message=FALSE, warning=TRUE}
library(GenomicSEM)
library(data.table)
library(ggplot2)
```

## Load datasets

```{r}
aid_sumstats <- fread('outputs/rev_1/02_munge_ldsc_sumstats/aid_sumstats.txt', data.table = F)
ldsc_model <- readRDS('outputs/rev_1/02_munge_ldsc_sumstats/ldsc_output_rev1.RDS')
factor_loci <-fread('outputs/rev_1/04_genomic_regions_factors/regions_factors.txt', data.table = F)

```

## Independet models estimation 

The independentcy is tested for the lead SNP of each significant genomic region. The lead SNP is the column 'SNP' in the factor_loci table. 
```{r}
loci_snps  <- aid_sumstats[which(aid_sumstats$SNP %in% factor_loci$SNP),]
```


### Model specification

One model per factor

```{r}
aid_model_f1 <- 'F1 =~ NA*crohn + uc  + psc  
F2 =~ NA*jia + sle + ra+ t1d 
F3 =~ NA*asthma+ derma 

F1~~F2
F1~~F3
F2~~F3
F1 ~~ 1*F1
F2 ~~ 1*F2
F3 ~~ 1*F3



F2 ~ SNP
F3 ~ SNP

crohn + uc  + psc ~ SNP


crohn~~a*crohn
uc~~b*uc
psc~~c*psc
jia~~d*jia
sle~~e*sle
ra~~f*ra
t1d~~g*t1d
asthma~~h*asthma
derma~~i*derma

a>0.001
b>0.001
c>0.001
d>0.001
e>0.001
f>0.001
g>0.001
h>0.001
i>0.001
'


aid_model_f2 <- 'F1 =~ NA*crohn + uc  + psc  
F2 =~ NA*jia + sle + ra+ t1d 
F3 =~ NA*asthma+ derma 

F1~~F2
F1~~F3
F2~~F3
F1 ~~ 1*F1
F2 ~~ 1*F2
F3 ~~ 1*F3



F1 ~ SNP
F3 ~ SNP

jia + sle + ra+ t1d ~ SNP

crohn~~a*crohn
uc~~b*uc
psc~~c*psc
jia~~d*jia
sle~~e*sle
ra~~f*ra
t1d~~g*t1d
asthma~~h*asthma
derma~~i*derma

a>0.001
b>0.001
c>0.001
d>0.001
e>0.001
f>0.001
g>0.001
h>0.001
i>0.001'


aid_model_f3 <- 'F1 =~ NA*crohn + uc  + psc  
F2 =~ NA*jia + sle + ra+ t1d 
F3 =~ NA*asthma+ derma 

F1~~F2
F1~~F3
F2~~F3
F1 ~~ 1*F1
F2 ~~ 1*F2
F3 ~~ 1*F3



F1 ~ SNP
F2 ~ SNP

asthma + derma~ SNP

crohn~~a*crohn
uc~~b*uc
psc~~c*psc
jia~~d*jia
sle~~e*sle
ra~~f*ra
t1d~~g*t1d
asthma~~h*asthma
derma~~i*derma

a>0.001
b>0.001
c>0.001
d>0.001
e>0.001
f>0.001
g>0.001
h>0.001
i>0.001'
```

### Estimation of independet models

It takes ~1 hour.  

```{r}
output <- list()

for(i in 1:3){
  tt <-  c('f1', 'f2', 'f3')[i]
  output[[tt]] <- userGWAS(covstruc = ldsc_model, 
                           SNPs = loci_snps, 
                           model = c(aid_model_f1, aid_model_f2, aid_model_f3)[i], 
                           sub=c('F1~~F2'), #the CHi square is the same for every 
                           parallel = T,cores = 2 )
}

saveRDS(output, 'outputs/rev_1/07_Qheterogeneity/output_model_estimation.RDS')

```
### Load factors summary statistics

```{r message=TRUE, warning=TRUE, paged.print=TRUE}
#load q het
q_ind <- readRDS('outputs/rev_1/07_Qheterogeneity/output_model_estimation.RDS')


#load factor summary stats
fac <- readRDS('outputs/rev_1/03_gwas_estimation/all_chunks_rev1.RDS')


#allocate the column of Q pvalue
factor_loci$Q_chisq_pval <- rep(NA, nrow(factor_loci))

```

### Calculate the Q pvalue 

Subtract the Chisquare of the independent model from the Chisquare of the primary model. Do the same for the degrees of freedom and calculate the pvalue of the resulting Chisquare. 

```{r}
#loop for each of the three factors
for(i in 1:3){
  
      #select the factor
      tt <- c('f1', 'f2', 'f3')[i]
      q_act <- q_ind[[tt]][[1]]
      #select the lead SNPs in heterogeneity sumstats for that factor
      q_act_2 <- q_act[which(q_act$SNP %in% factor_loci[which(factor_loci$trait==tt),]$SNP),]
      #select the lead SNPs in summary stats for that factor
      gwas <- fac[[1]][which(fac[[1]]$SNP %in% q_act_2$SNP), ]
      
      #put all of them in the same order
      gwas_2 <- gwas[match(q_act_2$SNP, gwas$SNP),]
      
      #substract the Chisquare
      Q <- gwas_2$chisq - q_act_2$chisq
      df <- gwas_2$chisq_df - q_act_2$chisq_df
      
      #put all of them in the same order
      factor_loci[which(factor_loci$trait==tt),][match(q_act_2$SNP, factor_loci[which(factor_loci$trait==tt),]$SNP, nomatch = NA),]$Q_chisq_pval<- pchisq(Q,df,lower.tail=FALSE)
      print(table(pchisq(Q,df,lower.tail=FALSE)<5e-8))
    
  }

#add a column stating wheter the Q pval is significant (TRUE)
factor_loci$het<- ifelse(factor_loci$Q_chisq_pval<5e-8, TRUE, FALSE)

#save
fwrite(factor_loci, 'outputs/rev_1/07_Qheterogeneity/genomic_regions_table_heterogeneity_column.txt', sep = '\t', col.names = T)

```


### Plot

```{r}

#save
pdf('outputs/rev_1/07_Qheterogeneity/heterogeneity_index_genomic_regions.pdf', width = 10, height = 10)
ggplot(data=factor_loci, aes(x=trait, fill=het) ) +
  geom_bar(stat='count',position = position_stack(reverse=T),  color='black')+
  scale_fill_manual(values = c("grey80", "white")) +
  geom_text(aes(label = paste0("n=", ..count..)),position= position_stack(vjust = 0.5, reverse=T),stat='count')+
  labs(y = 'Number of genomic regions', x = '')+
  theme_classic() +
  theme(legend.position="bottom") + ggtitle('Q heterogeneoity index lead SNPs genomic regions')
dev.off()

#plot
ggplot(data=factor_loci, aes(x=trait, fill=het) ) +
  geom_bar(stat='count',position = position_stack(reverse=T),  color='black')+
  scale_fill_manual(values = c("grey80", "white")) +
  geom_text(aes(label = paste0("n=", ..count..)),position= position_stack(vjust = 0.5, reverse=T),stat='count')+
  labs(y = 'Number of genomic regions', x = '')+
  theme_classic() +
  theme(legend.position="bottom") + ggtitle('Q heterogeneoity index lead SNPs genomic regions')

```



```{r}
sessionInfo()
```


